---
title: "Building a Predictive Model for NFL Offensive Playcalling"
format: pdf
editor: visual

execute:
  echo: false
  warning: false
  message: false
---

**Authors:** Jason Lin, Ryan Ji, Matthew Goetting, Patrick Shukis

## Introduction

Our dataset contains play-by-play data of every NFL play, from 2000 to 2024. The data is cited from official NFL-licensed publicly available information, aggregated to the readable csv files in the NFLFastR package.

Some of the main variables we plan to consider are the current down and distance, the position on the field, the type of play executed, win probability added (WPA), yards gained, and various others. Many of these metrics are explained in the introduction of the graph, and analysis is given to help interpret the graph for useful information.

Not only do we plot based upon numeric data, we also plot based on varying categorical data. For instance, graphing the number of plays executed by the differing types of plays throughout each NFL Division, for which there are 8. This allows for a much more immersive visual models that allows for better understanding of data and a higher interpretation of play-by-play offensive strategy in the NFL.

Furthermore, we have a large range of graphs utilized, from heat maps showcasing positional data, bar graphs demonstrating categorical data, to even a linear regression model of Go-For_It Decisions based on various in-game metrics.

## Research Question

Our main research objective is to create a predictive model for NFL offensive strategies. Based on the score, time left in the game, down and distance, position in the field, and more, can we predict which action an NFL team will take? Will they go for it on 4th and 2? Will they pass or run on 3rd and 4? For every situation, can we derive an "optimal" play choice? Through plotting, we are able to understand patterns in team play and intricacies among various factors like conference division, year, and individual team play. Our goal in this fourth draft is to explore different facets of NFL play-calling and identify patterns among hundreds of thousands of data entries.

With the implementation of SQL as the primary tool of data filtration, dataset indexing, aggregating, and joining is much faster and efficient. This opens up a variety of research opportunities that previously would have used a significant amount of computational complexity.

```{r}
library(nflfastR)
library(dplyr)
library(purrr)
library(DBI)
library(RSQLite)
library(readr)

years <- 2000:2024  
dcon <- dbConnect(RSQLite::SQLite(), "nfl_pbp.sqlite")

# creates tables for specific years (2000-2024)
for(year in years){
  message("Loading ", year, "...")
  pbp <- nflfastR::load_pbp(year)
  table_name <- paste0("pbp_", year)
  
  dbWriteTable(dcon, table_name, pbp, overwrite = TRUE)
  
  message("Created Table: ", table_name)
}

# creates the general play_by_play data from all years
all_queries <- paste0("SELECT * FROM pbp_", years)
union_query <- paste(all_queries, collapse = " UNION ALL ")

dbExecute(dcon, paste0("CREATE TABLE play_by_play AS ", union_query))

```

**Text Mining**

Here, we text mine to discover patterns in the descriptions given by the play-by-play data. We search for keywords within the datasets and return the number of times that word was discovered in the table. We look at three specific sections of time: all-time (2000-2024), only 2000, and only 2024. This is to help us with analyzing patterns in our graphs shown later.

```{r}
library(gt)

word_query <- "
SELECT 'all_time' AS season, 
       SUM(CASE WHEN \"desc\" LIKE '%pass%' THEN 1 ELSE 0 END) AS pass,
       SUM(CASE WHEN \"desc\" LIKE '%run%' THEN 1 ELSE 0 END) AS run,
       SUM(CASE WHEN \"desc\" LIKE '%touchdown%' THEN 1 ELSE 0 END) AS touchdown,
       SUM(CASE WHEN \"desc\" LIKE '%yard%' THEN 1 ELSE 0 END) AS yard,
       SUM(CASE WHEN \"desc\" LIKE '%down%' THEN 1 ELSE 0 END) AS down
FROM play_by_play

UNION ALL

SELECT '2000' AS season, 
       SUM(CASE WHEN \"desc\" LIKE '%pass%' THEN 1 ELSE 0 END) AS pass,
       SUM(CASE WHEN \"desc\" LIKE '%run%' THEN 1 ELSE 0 END) AS run,
       SUM(CASE WHEN \"desc\" LIKE '%touchdown%' THEN 1 ELSE 0 END) AS touchdown,
       SUM(CASE WHEN \"desc\" LIKE '%yard%' THEN 1 ELSE 0 END) AS yard,
       SUM(CASE WHEN \"desc\" LIKE '%down%' THEN 1 ELSE 0 END) AS down
FROM pbp_2000

UNION ALL

SELECT '2024' AS season, 
       SUM(CASE WHEN \"desc\" LIKE '%pass%' THEN 1 ELSE 0 END) AS pass,
       SUM(CASE WHEN \"desc\" LIKE '%run%' THEN 1 ELSE 0 END) AS run,
       SUM(CASE WHEN \"desc\" LIKE '%touchdown%' THEN 1 ELSE 0 END) AS touchdown,
       SUM(CASE WHEN \"desc\" LIKE '%yard%' THEN 1 ELSE 0 END) AS yard,
       SUM(CASE WHEN \"desc\" LIKE '%down%' THEN 1 ELSE 0 END) AS down
FROM pbp_2024
"

mining_table <- dbGetQuery(dcon, word_query)

mining_table %>%
  gt() %>%
  tab_header(
    title = "Word Frequency by Season"
  )

```

Here, we are able to specifically see different key words that are found in the descriptions of selected time periods in the NFL. We can see that "yard" is the most commonly found keyword, which is understandable because it is the main metric of distance used in football. Furthermore, we can see interesting trends between time periods. The word "touchdown" was utilized more often in descriptions of specific plays, which can attest to the NFL's increasingly more offensive-focused strategy between 2000 and 2024. The word "pass" is understandably the most used term, but it is surprising that "run" is so uncommon in the 2024 season descriptions, as the 2000 season writes it almost 1000 times more.

**Plot 1 (Scatterplot):**

Our first plot is more of an introduction to how we discover patterns between varying metrics in play-by-play data.

This scatter plot shows two lines: the blue is teams’ willingness to attempt 4th downs at each distance from the first-down marker, and the red line shows conversion success when they do. Notice that the blue line is not linear but exponential decay. 4th and 1 gets attempted \~48% of the time, but 4th and 2 drops to \~22%, which is a massive difference for just one yard. It keeps dropping from there: 4th and 5 is \~11% and everything past that decreases except for a sudden uptick at 10+ yards. This is selection bias because a lot of 4th and 10+ situations come from three straight incompletions, which happen when you’re trailing desperately. The red line (success rate) similarly follows the exponential pattern: each yard gained makes it exponentially easier to convert the 4th down. 

```{r}
library(ggplot2)
nfl_theme <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

pbp_4th <- pbp %>%
  filter(down == 4, !is.na(ydstogo), play_type %in% c("run", "pass", "punt", "field_goal")) %>%
  mutate(
    went_for_it = play_type %in% c("run", "pass"),
    success = went_for_it & yards_gained >= ydstogo,
    ytg_cat = case_when(
      ydstogo >= 10 ~ "10+",
      TRUE ~ as.character(ydstogo)
    ),
    ytg_cat = factor(ytg_cat, levels = c(as.character(1:9), "10+")),
    ydstogo_capped = pmin(ydstogo, 10)
  )

ytg_summary <- pbp_4th %>%
  group_by(ytg_cat) %>%
  summarise(
    go_rate = mean(went_for_it, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ytg_long <- ytg_summary %>%
  select(ytg_cat, go_rate, success_rate) %>%
  tidyr::pivot_longer(cols = -ytg_cat, names_to = "metric", values_to = "value")

ggplot(ytg_long, aes(x = ytg_cat, y = value, color = metric, group = metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3, aes(shape = metric)) +
  scale_color_manual(
    values = c("go_rate" = "blue", "success_rate" = "red"),
    labels = c("Go-for-it rate", "Success rate")
  ) +
  scale_shape_manual(
    values = c(19, 17),
    labels = c("Go-for-it rate", "Success rate")
  ) +
  labs(
    x = "Yards to go",
    y = "Rate",
    title = "Fourth Down Go-For-It Rate vs Success",
    color = NULL,
    shape = NULL
  ) +
  nfl_theme

```

The takeaway is that if we want to regress yards-to-go against go-for-it rate, we can’t pick a linear predictor.

With our text mining frequency table, we know that the term "down" is found at \~50000 times, while the term "yard" is found at a disproportionate \~870000 times throughout all NFL descriptions 2000-2024. This makes sense, as the table demonstrates that there is a huge amount of data for "yards to go" for specifically the fourth down go-for-it rate. Furthermore, "yards" is used as a general metric, which also explains the disproportionality.

**Plot 2 (Heatmap):**

With respect to our first graph, we also wanted to visualize where on the field these plays normally occur. A heat-map is perfect in demonstrating concentration along a multitude of metrics.

This heat map shows go-for-it rate by both yards-to-go (x-axis) and field position (y-axis, measured in yards from the opponent's goal line). The more yellow the color the more aggressive teams play. The most aggressive zone isn’t the red zone like you’d expect: it’s 21-40 yards from the opponent's goal. This is known as the “no man’s land” or “dead zone” in football, where you’re too far for a reliable field goal but close enough that punting does not change the opponent’s field position too much. The 0-20 (actual red zone) and 41-60 yard range come second. If you’re in the red zone, there's a lower risk to go for it because the opponent wouldn’t get a great position, but oftentimes you have an opportunity cost of 3 points. You rarely see teams go for it in the 81-100 range unless they’re truly desperate and behind several scores, because you risk giving the opponent the ball in your end zone. 

```{r}
obs_binned <- pbp_4th %>%
  mutate(
    bin = cut(yardline_100,
              breaks = c(0, 20, 40, 60, 80, 100),
              labels = c("0-20", "21-40", "41-60", "61-80", "81-100"),
              include.lowest = TRUE)
  ) %>%
  group_by(ytg_cat, bin) %>%
  summarise(went_for_it = mean(went_for_it, na.rm = TRUE),
            n = n(),
            .groups = "drop")

ggplot(obs_binned, aes(ytg_cat, bin, fill = went_for_it)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Go rate", limits = c(0, 1), trans = "sqrt") +
  labs(
    x = "Yards to go",
    y = "Field position (yards from opponent goal)",
    title = "Go-For-It Rate by Field Position & Distance",
  ) +
  nfl_theme +
  theme(legend.position = "right")

```

But across all distances except 21-40, we see that teams rarely go for it if it’s beyond 4th and 2. Only in the 21-40 range do you see teams often go for it in the 4th and 5 range. 

**Plot 3 (Histogram):**\
This graph shows how much yardage NFL teams gain on passing plays between 2000 and 2024. Each bar represents how frequently a pass play gained a certain number of yards, with the x-axis showing yardage gained and the y-axis showing the share of all plays. The dashed red and green lines mark the mean and median, which indicate the “average” and the “middle” value of all passing plays. For readers less familiar with football: a passing play is when the quarterback throws the ball forward to a receiver, and the number of “yards gained” is how far the ball advances from the line of scrimmage.

```{r}
library(dplyr)
library(ggplot2)
library(scales)

query <- "
SELECT 
    CASE 
        WHEN passing_yards > 60 THEN 60 
        ELSE passing_yards 
    END AS passing_yards
FROM play_by_play
WHERE passing_yards IS NOT NULL
"

pbp_pass <- dbGetQuery(dcon, query)
mean_y <- mean(pbp_pass$passing_yards, na.rm = TRUE)
med_y  <- median(pbp_pass$passing_yards, na.rm = TRUE)

ggplot(pbp_pass, aes(x = passing_yards)) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    bins = 30,
    fill = "#60a5fa", color = "white", alpha = 0.85
  ) +
  geom_vline(aes(xintercept = mean_y), color = "#ef4444", linewidth = 1, linetype = "dashed") +
  geom_vline(aes(xintercept = med_y), color = "#16a34a", linewidth = 1, linetype = "dotdash") +
  annotate("text", x = mean_y + 1, y = 0.09, label = paste0("Mean = ", round(mean_y, 1)), color = "#ef4444", hjust = 0, size = 3.2) +
  annotate("text", x = med_y + 1, y = 0.08, label = paste0("Median = ", round(med_y, 1)), color = "#16a34a", hjust = 0, size = 3.2) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = c(0, 0.002)) +
  scale_x_continuous(breaks = seq(-10, 60, by = 10), limits = c(-10, 60)) +
  labs(
    title = "Distribution of Passing Yards per Play, 2000–2024",
    x = "Passing Yards (capped at 60)",
    y = "Share of Plays",
    caption = "Dashed = Mean, Dot-Dash = Median"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(margin = margin(b = 5)),
    axis.text = element_text(color = "gray20")
  )

```

The chart highlights that most NFL passes are short or moderate in length.The right-skewed shape means that although a few deep passes can gain 30 or more yards, they are rare. The mean of 11.3 yards sits to the right of the median (9 yards), showing that occasional long completions pull the average upward. Overall, the figure illustrates that modern passing strategy relies heavily on short, efficient throws to sustain drives, with explosive plays serving as occasional outliers rather than the norm.

**Plot 4 (Line Graph):**

This chart examines how often NFL teams convert on third down based on the distance they need to gain, comparing running plays and passing plays. Third downs are the most critical moments because if a team gains enough yards for a first down, it extends the drive; if not, the drive usually ends. Each point shows the average conversion probability at a given yardage, and the dotted vertical lines mark common coaching thresholds at 3, 5, and 10 yards because those are situations where play-calling tendencies typically shift.

```{r}
library(dplyr)
library(ggplot2)
library(scales)

query <- "
SELECT
  ydstogo,
  play_type,
  COUNT(*) AS n,
  AVG(CASE WHEN first_down = 1 THEN 1 ELSE 0 END) AS conv_rate
FROM play_by_play
WHERE down = 3
  AND play_type IN ('run','pass')
  AND ydstogo BETWEEN 1 AND 15
  AND first_down IS NOT NULL
GROUP BY ydstogo, play_type
HAVING COUNT(*) >= 150
ORDER BY ydstogo, play_type
"

conv_tbl <- dbGetQuery(dcon, query)

# Plot
ggplot(conv_tbl, aes(x = ydstogo, y = conv_rate, color = play_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = c(3, 5, 10), linetype = "dotted", color = "gray55") +
  annotate("text", x = 3,  y = 0.98, label = "3 yds",  vjust = 0, size = 3, color = "gray35") +
  annotate("text", x = 5,  y = 0.98, label = "5 yds",  vjust = 0, size = 3, color = "gray35") +
  annotate("text", x = 10, y = 0.98, label = "10 yds", vjust = 0, size = 3, color = "gray35") +
  scale_y_continuous("3rd-Down Conversion Probability",
                     labels = percent_format(accuracy = 1),
                     limits = c(0, 1),
                     expand = expansion(mult = c(0, .06))) +
  scale_x_continuous("Yards to Go (3rd down)", breaks = seq(1, 15, 2)) +
  scale_color_manual(values = c(run = "#1f78b4", pass = "#33a02c"),
                     name = "Play Type",
                     labels = c("Run", "Pass")) +
  labs(
    title = "3rd-Down Conversions by Distance to Go: Run vs Pass",
    subtitle = "NFL plays, 2000–2024. Points shown only where sample size ≥ 150.",
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

Overall, conversion probability declines as the distance to go increases. At 1 yard, passes slightly outperform runs which is likely because teams use quick, high-percentage throws or play-action when defenses stack the line against the run. Between 5 and 7 yards, passing and running have basically identical percentages. However, beyond 10 yards, the data shows runs occasionally have a higher observed conversion rate. This isn’t because running is suddenly better—it’s because of selection bias. Teams almost never run on 3rd-and-long unless the defense is expecting a pass which may make a run safer. Those rare runs often catch defenses off guard and succeed at a higher rate, but they represent a very small portion of total plays.

**Plot 5 (Linear Regression Model):**

Now we built a logistic regression with 4 predictors: yards-to-go, field position, minutes remaining, and score differential. The coefficient plot shows the effect size of each factor. Yards-to-go is the most dominant factor: each additional yard reduces your odds of going for it by \~25%. The score differential comes second: teams that are behind get more aggressive. Minutes remaining is third: less time means more desperation. And field position has the least impact.

```{r}
model_final <- glm(
  went_for_it ~ (ydstogo_capped) + yardline_100 + 
    I(game_seconds_remaining/60) + score_differential,
  data = pbp_4th,
  family = binomial
)

coef_df <- broom::tidy(model_final, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      grepl("ytg_cat", term) ~ gsub("ytg_cat", "Yards to go: ", term),
      term == "yardline_100" ~ "Field position (Yards from Opposing Endzone)",
      term == "I(game_seconds_remaining/60)" ~ "Minutes Remaining",
      term == "score_differential" ~ "Score Differential",
      term == "ydstogo_capped" ~ "Yards to Go",
      TRUE ~ term
    )
  )

ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    x = "",
    y = "Coefficient (log odds)",
    title = "Which Factors Impact Go-For-It Decisions?",
  ) +
  nfl_theme

pbp_4th <- pbp_4th %>%
  mutate(
    pred_prob = predict(model_final, type = "response"),
    pred_decision = pred_prob > 0.5  # lowered from 0.5
  )

accuracy_stats <- pbp_4th %>%
  summarise(
    accuracy = mean(pred_decision == went_for_it),
    precision = sum(pred_decision & went_for_it) / sum(pred_decision),
    recall = sum(pred_decision & went_for_it) / sum(went_for_it)
  )

cat("\n=== Model Performance ===\n")
cat("Overall Accuracy:", round(accuracy_stats$accuracy * 100, 1), "%\n")
cat("Precision (when model says 'go', % correct):", round(accuracy_stats$precision * 100, 1), "%\n")
cat("Recall (% of actual go decisions caught):", round(accuracy_stats$recall * 100, 1), "%\n")

pbp_4th <- pbp_4th %>%
  mutate(
    error_type = case_when(
      pred_decision & !went_for_it ~ "false_positive",
      !pred_decision & went_for_it ~ "false_negative",
      TRUE ~ "correct"
    )
  )

cat("\n=== ERROR ANALYSIS ===\n\n")

fn <- pbp_4th %>% filter(error_type == "false_negative")
cat("FALSE NEGATIVES (we said punt, they went for it):\n")
cat("  Count:", nrow(fn), "\n")
cat("  Avg/Med yards to go:", round(mean(fn$ydstogo_capped), 1), "/", median(fn$ydstogo_capped), "\n")
cat("  Avg/Med field position:", round(mean(fn$yardline_100), 1), "/", median(fn$yardline_100), "\n")
cat("  Avg/Med score diff:", round(mean(fn$score_differential), 1), "/", median(fn$score_differential), "\n")
cat("  Avg/Med minutes left:", round(mean(fn$game_seconds_remaining/60), 1), "/", round(median(fn$game_seconds_remaining/60), 1), "\n")
cat("  % in final 5min:", round(mean(fn$game_seconds_remaining < 300) * 100, 1), "%\n\n")

fp <- pbp_4th %>% filter(error_type == "false_positive")
cat("FALSE POSITIVES (we said go, they punted):\n")
cat("  Count:", nrow(fp), "\n")
cat("  Avg/Med yards to go:", round(mean(fp$ydstogo_capped), 1), "/", median(fp$ydstogo_capped), "\n")
cat("  Avg/Med field position:", round(mean(fp$yardline_100), 1), "/", median(fp$yardline_100), "\n")
cat("  Avg/Med score diff:", round(mean(fp$score_differential), 1), "/", median(fp$score_differential), "\n")
cat("  Avg/Med minutes left:", round(mean(fp$game_seconds_remaining/60), 1), "/", round(median(fp$game_seconds_remaining/60), 1), "\n")
cat("  % in final 5min:", round(mean(fp$game_seconds_remaining < 300) * 100, 1), "%\n\n")

```

The results of the model on the full dataset are mixed. The 87.9% accuracy sounds high, but it is somewhat misleading when you realize that lots of situations are “obvious”, such as punts or field goals, which occur most of the time. The recall at 31.2% percent tells us that we’re missing 70% of the actual go-for-it decisions, which tells us that the model is too conservative. The precision is 65.6%, which tells us that when we do predict go-for-it rate, we’re right ⅔ of the time.

We conducted an error analysis to see the cases our model missed. There were 9700 cases where the model predicted punt/kick but they went for it. The average false negative situation happens at midfield (opposing 40 yd line), a team slightly behind (-3 points), normal game time left (23 min). This is where aggressive coaches might make the decision but our model is trained on league average behavior (and misses the increase in go-for-it rate in recent years). 

There were 2309 cases where our model predicted that teams would go for it but they didn’t. The average false positive situation happened at the opposing 25 yd line, down by 10, with 12 min left on the clock with 3 yards to go. So the model might read desperation (down by 10 and 4th quarter) and a short yardage situation, but in reality the team might opt to take the field goal to cut the lead to one score. 

Overall the model is relatively accurate, but misses the recent uptick in go-for-it rate and individual play-calling habits of coaches. The false positives and negatives usually happen in borderline cases where you could make a reasonable argument either way.

**Plot 6 (Bar Graph):**

All of the previous graphs give us a general understanding of offensive strategy and metrics relate to each other in the NFL. However, from here, we want to shift toward understanding the modern NFL.

This chart compares the average touchdown distance for running and passing plays in the 2000 and 2024 NFL seasons. Each bar represents the mean yardage gained on touchdowns of that type, allowing for a side-by-side look at how scoring play dynamics have evolved over two decades. The goal is to understand whether modern offenses are producing shorter or longer touchdown plays, and how the balance between passing and rushing has changed over time.

```{r}
library(dplyr)
library(ggplot2)

query <- "
SELECT
    season,
    CASE play_type
        WHEN 'run' THEN 'Run'
        WHEN 'pass' THEN 'Pass'
    END AS play_type,
    AVG(yards_gained) AS avg_len
FROM (
    SELECT *, 2000 AS season FROM pbp_2000
    UNION ALL
    SELECT *, 2024 AS season FROM pbp_2024
) AS combined
WHERE td_team IS NOT NULL
  AND play_type IN ('run', 'pass')
GROUP BY season, play_type
ORDER BY season, play_type
"

td_tbl <- dbGetQuery(dcon, query)


# Soft, desaturated colors
cols <- c("Run" = "#5E81AC", "Pass" = "#A3BE8C")

ggplot(td_tbl, aes(x = factor(season), y = avg_len, fill = play_type)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5, color = NA) +
  scale_fill_manual(values = cols, name = "Play Type") +
  coord_cartesian(ylim = c(0, max(td_tbl$avg_len, na.rm = TRUE) + 3)) +
  labs(
    title = "Average Touchdown Distance by Play Type",
    subtitle = "NFL seasons 2000 and 2024",
    x = "Season",
    y = "Average Touchdown Distance (yards)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )



```

The results show that passing touchdowns remain significantly longer than rushing touchdowns in both eras, although both types have shortened slightly since 2000. In 2000, passes averaged around twice the distance of runs, suggesting that deep throws played a larger role in scoring. By 2024, average touchdown distances declined for passing likely reflecting more conservative offensive schemes and improved defensive coverage in the modern game. Still, the relative gap persists—indicating that while football strategy has evolved, the pass continues to drive longer, more explosive scoring plays than the run.

Interestingly enough, our text mining data shows the opposite trend. The term "run" appears almost 1000 times less in the 2024 season compared to the 2000 season, however, this graph shows that runs as a play are used more often in 2024 than 2000. This variability could be attested as simply the 2000 season described plays differently, or possibly a trend in how runs differ in 2024 compared to 2000.

**Plot 7 (Layered Histogram):**

Along with running and passing, there are other metrics available that allows us to understand the modern NFL's offensive play-by-play strategy. Even though it is important to understand how NFL teams operate, it is also imperative to understand the impact each of these plays have on the outcome of the game.

Win Probability Added, or WPA, is an advanced statistic that measures the impact that a player has to the victory of a game through the effect that a certain play made by the player had on the outcome of the game. Furthermore, I define clutch time as any entry of a play that is during the last 60 seconds of a game with a score differential of less than or equal to 8 points. This is because there is still an opportunity for the losing team to win during this time through a single play, or a touchdown and a two-point conversion. Lastly, success in the NFL is defined by gaining a certain percentage of yardage by the offense, for instance, 50% or more of the needed yards in the 1st down.

```{r}
library(stringr)
library(nflreadr)
library(nflfastR)
library(nflplotR)
library(ggplot2)
library(dplyr)

query <- "
SELECT *
FROM play_by_play
WHERE season = 2024
"

#df <- dbGetQuery(dcon, query)
#head(df)

rosters <- nflreadr::load_rosters(2000:2024)

qb_query <- "
SELECT
    passer_player_name,
    season,
    SUM(passing_yards) as total_pass_distance,
    AVG(passing_yards) as avg_pass_distance
FROM play_by_play
WHERE passer_player_name IS NOT NULL
  AND passing_yards IS NOT NULL
  AND (season = 2000 OR season = 2024)
GROUP BY 
        passer_player_name,
        season
ORDER BY 
        season,
        total_pass_distance DESC,
        avg_pass_distance DESC,
        passer_player_name
"

qb_df <- dbGetQuery(dcon, qb_query)
qb_df

top_qbs_2000 <- qb_df %>% filter(season == 2000) %>% head(20)
top_qbs_2024 <- qb_df %>% filter(season == 2024) %>% head(20)

top_qbs <- bind_rows(top_qbs_2000, top_qbs_2024)
top_qbs


# Use the combined table
plot_df <- top_qbs %>%
  mutate(
    season = factor(season),       # ensures color groups
    passer_player_name = as.factor(passer_player_name)
  )

ggplot(plot_df, aes(x = total_pass_distance,
                    y = avg_pass_distance,
                    color = season,
                    label = passer_player_name)) +

  geom_point(size = 4, alpha = 0.8) +

  # Label QBs
  ggrepel::geom_text_repel(size = 2, max.overlaps = 20) +

  scale_color_manual(
    values = c("2000" = "blue", "2024" = "red")
  ) +

  labs(
    title = "QB Passing Profiles: Total vs Average Yards (2000 vs 2024)",
    x = "Total Passing Yards",
    y = "Average Yards per Attempt",
    color = "Season"
  ) +

  theme_minimal(base_size = 14)

```

The graph indicates that in 2024, the win probability added by successful plays compared to failed plays during the most crucial points in the game average out at around 0% probability incremented, however, successful plays tend to have increasingly more positive WPA, while failed plays tend to bulk toward being harmful to the team’s probability of victory. Certain successful plays even reach around 80% win probability added, while certain failed plays can be detrimental, sometimes decreasing a team’s win probability by \~70%. Using this graph, we are able to derive how substantial of a difference the impact a successful play has over a failed play throughout 2024.

**Plot 8 (Facet-Wrapped Bar Graph):**

Not only are league-wide statistics important to analyze to understand the modern NFL, other patterns may arise from subsetting the data into smaller factors to understand strategy on a smaller scale. For instance, there may be similarities to how certain divisions play in comparison to others.

The context of this graph is that I wanted to showcase the differences in play type ran by division, as I wanted to see differences between each division in the NFL. The play types recognized in nflfastR are categorized into 9 different types of plays, including "No Play", which indicates a play was not executed. I took out the x labels because I have the total Number of Plays labelled on each bar, allowing for a much more visually appealing graph. Furthermore, I chose the 2024 season because it is the most contemporary data available, so these play types are as relevant as it is feasible to analyze. Lastly, I chose a bar graph to demonstrate the differences in the number of plays executed by each play type, as a bar graph here is best.

```{r}
library(stringr)
library(nflreadr)
library(ggplot2)
library(dplyr)

query <- "
SELECT
  season,
  SUM(CASE WHEN extra_point_attempt = 1 THEN 1 ELSE 0 END) AS xp_attempts,
  SUM(CASE WHEN extra_point_attempt = 1 AND extra_point_result = 'good' THEN 1 ELSE 0 END) AS xp_made
FROM play_by_play
GROUP BY season
HAVING season BETWEEN 2000 AND 2024
ORDER BY season
"

xp_df <- dbGetQuery(dcon, query) %>% mutate(xp_pct = xp_made/xp_attempts)
mean_xp_before <- mean(xp_df$xp_pct[xp_df$season >= 2000 & xp_df$season <= 2014], na.rm = TRUE)
mean_xp_after <- mean(xp_df$xp_pct[xp_df$season >= 2015], na.rm = TRUE)

ggplot(xp_df, aes(x = season, y = xp_pct)) +
  geom_line(size = 1.2, color = "mediumorchid3") +
  geom_point(size = 3, color = "mediumpurple4") +
  
  geom_segment(
    aes(x = 2000, xend = 2014, y = mean_xp_before, yend = mean_xp_before),
    color = "red",
    linetype = "dashed",
    linewidth = 1) +
  annotate("text",
           x = min(xp_df$season) + 5,
           y = mean_xp_before + 0.002,
           label = paste0("Mean: ", scales::percent(mean_xp_before, accuracy = 0.1)),
           vjust = -1,
           color = "red",
           size = 4) +
  
  geom_segment(
    aes(x = 2015, xend = 2024, y = mean_xp_after, yend = mean_xp_after),
    color = "blue",
    linetype = "dashed",
    linewidth = 1) +
  annotate("text",
           x = max(xp_df$season) - 5,
           y = mean_xp_after + 0.002,
           label = paste0("Mean: ", scales::percent(mean_xp_after, accuracy = 0.1)),
           vjust = -1,
           color = "blue",
           size = 4) +
  
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 2)) +
  labs(
    title = "NFL Extra Point Conversion Rate by Season (2000–2024)",
    x = "Season",
    y = "Extra Point Success Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )


```

As seen above, we can see that the most common play throughout the NFL in all division are pass plays, which makes sense because passes are the focal point of offensive strategy in the NFL, followed by runs. The rest of the plays are significantly less utilized, however, we can still find differences between play types in differing conferences. For instance, the NFC East and AFC West, having around 2600 pass plays, pass way more than the NFC South, only having around 2400. Furthermore, the AFC West QB Kneel much more often than the AFC South, from 77 to 36. With this graph, we are able to understand the differences in plays ran by conference division, allowing us to notice patterns in offensive strategies within certain divisions or a general pattern in the entire league.

In our text mining demonstration, we can see that the pass play was very obviously the most common play in 2024. However, runs are represented at a much larger proportion compared to their descriptions. This is likely due to passes being described directly while the descriptions of runs do not include the term "run" as often. However, text mining supports our understanding of modern NFL strategy through demonstrating how passes are the most common offensive play.

**Plot 9 (Facet-Wrapped Hex Heat Map):**

Going along with divisional data, we can analyze specific offensive metrics as well to see patterns between conference divisional offensive strategy.

For some context, air yards refers to the distance a football travels in the air from the line of scrimmage to the receiver’s location, while yards after catch measures the distance a receiver travels during a play after receiving the ball. I chose the 2024 season because it is contains the most contemporary, relevant data, allowing us to understand the scope of concurrent strategy in the NFL. Furthermore, the graph is divided into individual conference divisions, allowing us to visualize how each division differs in their offensive strategy and how they compare to other divisions. The hex map parallels a heat map, and allows the viewer to understand the relationship between air yardage and running yardage of each division.

```{r}
library(hexbin)
library(nflreadr)
library(ggplot2)
library(scales)

#teams <- load_teams()
#dbWriteTable(dcon, "teams", teams, overwrite = TRUE)

pct_query <- "
WITH filtered AS (
    SELECT *
    FROM play_by_play
    WHERE posteam IS NOT NULL
      AND air_yards IS NOT NULL
      AND yards_after_catch IS NOT NULL
),
season_summary AS (
    SELECT
        season,
        SUM(air_yards) AS total_air_yards,
        SUM(yards_after_catch) AS total_yac,
        SUM(air_yards) + SUM(yards_after_catch) AS total_yards
    FROM filtered
    GROUP BY season
)

SELECT season,
      'pct_air_yards' AS metric,
      (total_air_yards * 1.0 / total_yards) AS value
FROM season_summary
      
UNION ALL

SELECT season,
      'pct_yac' AS metric,
      (total_yac * 1.0 / total_yards) AS value
FROM season_summary
"

pct_df <- dbGetQuery(dcon, pct_query)

pct_df <- pct_df %>%
  transmute(
    season,
    value,
    metric = recode(metric,
                    "pct_yac" = "yac (%)",
                    "pct_air_yards" = "air yards (%)")
  )

pct_df$metric <- factor(pct_df$metric,
                        levels = c("yac (%)", "air yards (%)"))

ggplot(pct_df, aes(x = factor(season), y = value, fill = metric)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(value, accuracy = 0.1)),
            position = position_stack(vjust = 0.5),
            size = 3,
            color = "black") +
  labs(
    title = "Proportion of Air Yards and Yards After Catch, by Season",
    x = "Season",
    y = "Yards"
  ) +
  scale_fill_manual(values = c(
    "yac (%)" = "darkseagreen",
    "air yards (%)" = "turquoise4"
  )) + scale_x_discrete(drop = FALSE) + coord_flip() + theme_minimal()

```

It is important to measure these metrics because it provides information on the structure of a certain division’s offense: whether a division prefers further passes or shorter passes with longer rush yards, measured by yards after catch. We can see that all of the teams average most of their passes short with a low yards-after-catch, indicating that all of these teams usually throw short and the receivers are not able to run far afterwards. Furthermore, we can also see that the NFC South division do not have an air pass over 50 yards, which may be an indicator that this division's Quarterbacks prefer not to throw far, receivers find it difficult to find openings, or the NFC South, as a collective division, choose shorter, safer passes over long passes. This could be an indicator of some trend with this certain area as a whole rather than of specific teams or players.

**Plot 10 (Lollipop Graph):**

Going to further specifics, we can subset the data by team, allowing us to understand in further detail of modern NFL strategy and how different teams operate. The rest of the graphs are all team-based plots that we can compare side by side to see patterns on a micro scale.

For some context, 3rd Down Conversion Percentage represents the percentage of the time that the offensive team is able to obtain a first down while on third down. This is a vital down for teams as most of the time that teams fail to convert on third down, they are forced to punt the ball away and lose possession, thus ending the drive. Third down conversions keep drives going and allow good teams to come out ahead, as this statistic in particular is extremely reflective of the NFL's top teams and how they are able to perform on third down.

```{r}
library(ggplot2)
library(nflplotR)

query <- "
SELECT
  posteam,
  CAST(SUM(CASE WHEN first_down = 1 THEN 1 ELSE 0 END) AS FLOAT) /
    (SUM(CASE WHEN first_down = 0 THEN 1 ELSE 0 END) +
     SUM(CASE WHEN first_down = 1 THEN 1 ELSE 0 END)) AS third_conv
FROM pbp_2024
WHERE down = 3
  AND posteam IS NOT NULL
  AND first_down IS NOT NULL
GROUP BY posteam
ORDER BY third_conv DESC
"

thirds <- dbGetQuery(dcon, query)

thirds %>%
  ggplot(aes(x = posteam, y = third_conv)) + 
  geom_segment(aes(xend = posteam, y = 0, yend = third_conv)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .036) +
  labs(
    title = "2024 NFL Third Down Conversion Rates",
    x = NULL,
    y = "Third Down Conversion Percentage"
  ) + theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

               
```

From a statistical perspective, 11 out of the top 15 teams in the NFL made the playoffs in the 2024 season, cementing the importance of third down conversion rate in running a successful offense. Teams ought to push for over a 50% conversion rating, which is the highest obtained by the Tampa Bay Buccaneers, as doing so could create an even more effective offensive scheme.

**Plot 11 (Team-Comparison Scatterplot):**

For some context, the final five minutes are incredibly important for trailing teams from an offensive standpoint. Typically, teams run a surplus of passing plays compared to running plays during “clutch time”, but nonetheless, this graph separates out teams by their skill at each of the two. As the game nears closer and closer to the end, more passing plays are performed by the trailing team, but the use of run plays in this graph determines teams who excel at running the ball in order to set themselves up for taking deep shots downfield in order to come back. For this graph, I only used data from teams who were trailing, and plays that were in the final five minutes of the game. This prevented the data from being skewed by leading teams who only call short run plays to run out the clock and secure their victory.

```{r}
library(ggplot2)
library(nflplotR)
library(ggrepel)

query <- "
SELECT
  posteam,
  AVG(CASE WHEN play_type = 'run' THEN yards_gained END) AS rush_avg,
  AVG(CASE WHEN play_type = 'pass' THEN yards_gained END) AS pass_avg,
  SUM(CASE WHEN play_type = 'run' THEN 1 ELSE 0 END) AS rush_plays,
  SUM(CASE WHEN play_type = 'pass' THEN 1 ELSE 0 END) AS pass_plays,
  SUM(CASE WHEN complete_pass = 1 THEN 1 ELSE 0 END) * 1.0 / 
    NULLIF(SUM(CASE WHEN play_type = 'pass' THEN 1 ELSE 0 END), 0) AS comp_pct
FROM pbp_2024
WHERE play_type IN ('run','pass')
  AND game_seconds_remaining < 301
  AND score_differential < 0
  AND play_type IS NOT NULL
GROUP BY posteam
ORDER BY posteam
"

two_min <- dbGetQuery(dcon, query)

ggplot(two_min, aes(x = pass_avg, y = rush_avg)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .03) +
  geom_text_repel(aes(label = posteam), size = 2.5) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) + 
  labs(
    title = "Average Rush vs. Pass Yards by Trailing Teams \nin Final 5 Minutes",
    subtitle = "2024 NFL Season",
    x = "Average Pass Yards",
    y = "Average Rush Yards"
  )
```

In the graph, four teams in particular stand out in the graph. On the lower end, the Bills and the Chargers stand out from the pack with extremely low average pass yards per play. This is mainly due to their low completion percentages in clutch time, with the Chargers having a 29% completion percentage, and the Bills having a completion percentage of just 11%. On the other end of the spectrum we see the Packers, who excel in clutch time passing. They don’t have an abnormally high completion percentage, only 61% of passes completed, putting them near the top of the pack in this category, but not a clear frontrunner. The statistic is especially astounding because in order for them to achieve almost 10 yards per pass on only 61% completions, their average completion distance is over 15 yards, which is extremely successful in the final five minutes of games. The Ravens, who have a dual-pronged rushing attack, achieved the highest average rush yards by far, averaging over 10 yards on each rushing attempt, which is incredibly effective. They are able to do this with their all-pro running back Derrick Henry who averages a high yards after contact, and their mobile quarterback Lamar Jackson, who frequently gets out of the pocket and upfield for large rushing gains. The Ravens are also 3rd in average pass yards, making them arguably the most dangerous all-around team in clutch time, resulting in a few comeback wins last season.

**Plot 12 (Logo Scatterplot):**

For some context, the graph contains a comparison between percentage of passing plays called by offenses throughout the 2024 NFL Season and their total offensive yards. While passing plays are often higher yardage plays, it would lead one to believe that a team that passes more often would get more yardage. The argument can also be made that keeping a roughly even split is best for keeping the defense on their toes and remaining unpredictable. Based on this graph, there is a negative correlation between passing percentage and total yards.

```{r}
library(ggplot2)
library(nflplotR)

query <- "
SELECT
  posteam,
  COUNT(*) AS total_plays,
  SUM(yards_gained) AS total_yards,
  SUM(CASE WHEN play_type = 'pass' THEN 1 ELSE 0 END) * 100.0 / COUNT(*) AS pass_pct
FROM pbp_2024
WHERE play_type IN ('run','pass')
  AND play_type IS NOT NULL
GROUP BY posteam
ORDER BY posteam
"

pass_imp <- dbGetQuery(dcon, query)

ggplot(pass_imp, aes(x = pass_pct, y = total_yards)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .03) + geom_text_repel(aes(label = posteam), size = 2.5) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) + 
  labs(
    title = "Passing Percentage vs. Total Offensive Yards",
    subtitle = "2024 NFL Season",
    x = "Percentage of Passing Plays",
    y = "Total Offensive Yards"
  )

```

There are a couple possible reasons for this. One of these is that the teams on the lower end of the spectrum of passing percentage hover around the fifty percent mark, and even possibly below. Having an even distribution allows teams to be unpredictable and demonstrate that they can gain positive yardage with both the pass and the run. Teams who run more running plays have to be efficient in both the run and the pass, because if the run plays did not work efficiently, it would lead to a high rate of passing plays due to longer down and distance situations. Another explanation is the fact that winning teams that have more yards often control the game more and are in the lead. Because of this, they run more running plays to run the clock when they are ahead. On the other end of the spectrum, teams who trail more often run more passing plays in order to keep time on the clock and attempt more explosive plays for larger gains.
