---
title: "The Evolution of NFL Offensive Strategy"
subtitle: "Jason Lin, Matthew Goetting, Patrick Shukis, and Ryan Ji"
format:
  revealjs: 
    # embed-resources: true
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: styles.css
    footer: STAT 405 - Group 14
server: shiny
---
```{r}
#| echo: false
#| context: setup
library(nflfastR)
library(nflplotR)
library(nflreadr)
library(dplyr)
library(purrr)
library(DBI)
library(RSQLite)
library(readr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(tidyverse)
library(patchwork)
library(scales)
library(broom)

dcon <- dbConnect(RSQLite::SQLite(), "nfl_pbp.sqlite")

years <- 2000:2025

pbp <- map_df(years, function(y) {
  message("Loading ", y, "...")
  nflfastR::load_pbp(y)
})

pbp_2000 <- dbGetQuery(dcon, "SELECT * FROM pbp_2000")
pbp_2024 <- dbGetQuery(dcon, "SELECT * FROM pbp_2024")

nfl_theme <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```


## Background
::: incremental
-   All of us have interests corresponding with professional sports.
-   Of course, we wanted to cater our final project centering around sports statistics.
-   NFLFastR scrapes public NFL play-by-play data into a reliable R package
-   NFLFastR is an ideal foundation for studying how offensive strategy has evolved in the league
:::

## History
::: incremental
-   The early years of the league were built around the run game. 
-   Over time, the NFL introduced a series of rules that helped open up the field, protect receivers
and quarterbacks, and encourage more passing.
-   Kicker Accuracy improved over time along with various rule changes, affecting how kicking is utilized
in offensive strategy
-   As passing efficiency skyrocketed, offensive efficiency became more of a numbers game, prioritizing
positive value per play 
:::

## Our Mission

::: incremental
-   Understand the changes in offensive strategy in the time period between 2000 to 2025
-   Discern the effects of certain rule changes on NFL offense
-   Present our findings to NFL Executives, Teams, and Analysts
:::

## Approach
::: incremental
-	Use NFLFastR play-by-play data to build a consistent and reproducible dataset
-	Define key offensive metrics for comparison across eras
-	Create visualizations that highlight league-wide shifts in behavior rather than team-specific noise
-	Interpret results in the context of rule changes and broader analytical trends in football
:::

## Generating Our SQLite
```r
#| echo: true
years <- 2000:2025  
dcon <- dbConnect(RSQLite::SQLite(), "nfl_pbp.sqlite")

# creates tables for specific years (2000-2025)
for(year in years){
  message("Loading ", year, "...")
  pbp <- nflfastR::load_pbp(year)
  table_name <- paste0("pbp_", year)
  
  dbWriteTable(dcon, table_name, pbp, overwrite = TRUE)
  
  message("Created Table: ", table_name)
}

# creates the general play_by_play data from all years
all_queries <- paste0("SELECT * FROM pbp_", years)
union_query <- paste(all_queries, collapse = " UNION ALL ")

dbExecute(dcon, paste0("CREATE TABLE play_by_play AS ", union_query))

```

## Offensive Conversion
::: incremental
-    Drives now end in touchdowns and field goals more often than in the early 2000s
-    Red zone execution have increased across teams
-    Modern offenses waste fewer possessions through punts and turnovers
:::

## Changes in NFL Drives Between Eras (2000-2005 and 2020-2025)
```{r}
query_drive_change <- "
WITH period1 AS (
    SELECT fixed_drive_result, COUNT(*) as count_2000
    FROM (
        SELECT game_id, drive, fixed_drive_result
        FROM play_by_play
        WHERE season BETWEEN 2000 AND 2004
            AND fixed_drive_result NOT IN ('End of half', 'Safety', 'Opp touchdown')
        GROUP BY game_id, drive, fixed_drive_result
    )
    GROUP BY fixed_drive_result
),
period2 AS (
    SELECT fixed_drive_result, COUNT(*) as count_2020
    FROM (
        SELECT game_id, drive, fixed_drive_result
        FROM play_by_play
        WHERE season BETWEEN 2020 AND 2025
            AND fixed_drive_result NOT IN ('End of half', 'Safety', 'Opp touchdown')
        GROUP BY game_id, drive, fixed_drive_result
    )
    GROUP BY fixed_drive_result
)
SELECT 
    p1.fixed_drive_result,
    100.0 * p1.count_2000 / (SELECT SUM(count_2000) FROM period1) as pct_2000,
    100.0 * p2.count_2020 / (SELECT SUM(count_2020) FROM period2) as pct_2020,
    100.0 * p2.count_2020 / (SELECT SUM(count_2020) FROM period2) - 
    100.0 * p1.count_2000 / (SELECT SUM(count_2000) FROM period1) as diff
FROM period1 p1
JOIN period2 p2 ON p1.fixed_drive_result = p2.fixed_drive_result
ORDER BY ABS(diff) DESC
"


drive_change_data <- dbGetQuery(dcon, query_drive_change)

ggplot(drive_change_data, aes(x = reorder(fixed_drive_result, diff), y = diff, fill = diff > 0)) +
    geom_bar(stat = "identity", width = 0.7) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
    scale_fill_manual(
        values = c("FALSE" = "#E74C3C", "TRUE" = "#27AE60"),
        labels = c("Decreased", "Increased"),
        name = NULL
    ) +
    coord_flip() +
    labs(
        title = "How NFL Drives End Has Changed Dramatically",
        subtitle = "Percentage point change in drive outcomes (2020-2025 vs 2000-2005)",
        x = NULL,
        y = "Percentage Point Change"
    ) + nfl_theme


```

## Comparing Red Zone Touchdown Efficiency
```{r}
redzone_2024 <- pbp_2024 %>%
  filter(!is.na(posteam), yardline_100 <= 20, yardline_100 > 0) %>%
  group_by(posteam, drive) %>%
  slice(1) %>%
  ungroup() %>%
  group_by(posteam) %>%
  summarise(
    rz_trips = n(),
    rz_tds = sum(fixed_drive_result == "Touchdown", na.rm = TRUE),
    rz_td_pct = rz_tds / rz_trips
  )

# Get 2024 standings
standings_2024 <- pbp_2024 %>%
  filter(!is.na(posteam), !is.na(home_team), !is.na(away_team)) %>%
  group_by(game_id, posteam) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    win = case_when(
      posteam == home_team & result > 0 ~ 1,
      posteam == away_team & result < 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(posteam) %>%
  summarise(wins = sum(win)) %>%
  arrange(wins)

redzone_2024 <- redzone_2024 %>%
  left_join(standings_2024, by = "posteam") %>%
  arrange(wins) %>%
  mutate(posteam = factor(posteam, levels = posteam))

# 2000 Red Zone Efficiency
redzone_2000 <- pbp_2000 %>%
  filter(!is.na(posteam), yardline_100 <= 20, yardline_100 > 0) %>%
  group_by(posteam, drive) %>%
  slice(1) %>%
  ungroup() %>%
  group_by(posteam) %>%
  summarise(
    rz_trips = n(),
    rz_tds = sum(fixed_drive_result == "Touchdown", na.rm = TRUE),
    rz_td_pct = rz_tds / rz_trips
  )

# Get 2000 standings
standings_2000 <- pbp_2000 %>%
  filter(!is.na(posteam), !is.na(home_team), !is.na(away_team)) %>%
  group_by(game_id, posteam) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    win = case_when(
      posteam == home_team & result > 0 ~ 1,
      posteam == away_team & result < 0 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(posteam) %>%
  summarise(wins = sum(win)) %>%
  arrange(wins)

redzone_2000 <- redzone_2000 %>%
  left_join(standings_2000, by = "posteam") %>%
  arrange(wins) %>%
  mutate(posteam = factor(posteam, levels = posteam))

# Calculate averages
avg_2024 <- mean(redzone_2024$rz_td_pct, na.rm = TRUE)
avg_2000 <- mean(redzone_2000$rz_td_pct, na.rm = TRUE)

# Get combined y limits
y_limits_4 <- c(
  min(c(redzone_2024$rz_td_pct, redzone_2000$rz_td_pct), na.rm = TRUE),
  max(c(redzone_2024$rz_td_pct, redzone_2000$rz_td_pct), na.rm = TRUE)
)

# Create plots
g_2000_rz <- redzone_2000 %>%
  ggplot(aes(x = posteam, y = rz_td_pct)) + 
  geom_segment(aes(xend = posteam, y = 0, yend = rz_td_pct)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .036) +
  geom_hline(yintercept = avg_2000, linetype = "dotted", color = "red", linewidth = 0.8) +
  annotate("text", x = 1, y = avg_2000, label = sprintf("League Avg: %.1f%%", avg_2000 * 100), 
           hjust = 0, vjust = -0.5, color = "red", size = 3.5, fontface = "bold") +
  coord_cartesian(ylim = y_limits_4) +
  labs(
    title = "2000 Season",
    x = NULL,
    y = "Red Zone TD %"
  ) + 
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

g_2024_rz <- redzone_2024 %>%
  ggplot(aes(x = posteam, y = rz_td_pct)) + 
  geom_segment(aes(xend = posteam, y = 0, yend = rz_td_pct)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .036) +
  geom_hline(yintercept = avg_2024, linetype = "dotted", color = "red", linewidth = 0.8) +
  annotate("text", x = 1, y = avg_2024, label = sprintf("League Avg: %.1f%%", avg_2024 * 100), 
           hjust = 0, vjust = -0.5, color = "red", size = 3.5, fontface = "bold") +
  coord_cartesian(ylim = y_limits_4) +
  labs(
    title = "2024 Season",
    x = NULL,
    y = "Red Zone TD %"
  ) + 
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Combine plots
combined_plots_4 <- (g_2000_rz | g_2024_rz) +
  plot_annotation(
    title = "Red Zone Touchdown Efficiency",
    subtitle = "2000 vs. 2024 (Sorted by Wins)",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

combined_plots_4


```



## Kicking Throughout the NFL
::: incremental
-    Field Goals have become a staple of the modern NFL offensive game
-    Helps secure easier points when touchdowns aren't guaranteed
-    Gradual evolution throughout the years
-    Rule Changes have affected kicking instantaneously
:::

## Upward Long Range Field Goal Trends
```{r}
nfl_theme <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 11, hjust = 0, color = "gray40"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

query1 <- "
WITH long_fg_attempts AS (
    SELECT 
        season,
        COUNT(*) as fg_attempts_55plus
    FROM play_by_play
    WHERE play_type = 'field_goal' 
        AND kick_distance >= 55
    GROUP BY season
),
total_games AS (
    SELECT 
        season,
        COUNT(DISTINCT game_id) as total_games
    FROM play_by_play
    GROUP BY season
)
SELECT 
    t.season,
    COALESCE(l.fg_attempts_55plus, 0) as fg_attempts_55plus,
    t.total_games,
    CAST(COALESCE(l.fg_attempts_55plus, 0) AS FLOAT) / t.total_games as attempts_per_game
FROM total_games t
LEFT JOIN long_fg_attempts l ON t.season = l.season
ORDER BY t.season
"

fg_attempts_data <- dbGetQuery(dcon, query1)


query2 <- "
SELECT 
    season,
    COUNT(*) as total_attempts,
    SUM(CASE WHEN field_goal_result = 'made' THEN 1 ELSE 0 END) as made,
    CAST(SUM(CASE WHEN field_goal_result = 'made' THEN 1 ELSE 0 END) AS FLOAT) / COUNT(*) * 100 as success_rate
FROM play_by_play
WHERE play_type = 'field_goal' 
    AND kick_distance >= 55
GROUP BY season
ORDER BY season
"

fg_success_data <- dbGetQuery(dcon, query2)
combined_data <- merge(fg_attempts_data, fg_success_data, by = "season")

# calculate scaling factor
coeff <- 100 / max(combined_data$attempts_per_game)

ggplot(combined_data, aes(x = season)) +
    geom_line(aes(y = attempts_per_game, color = "Attempts per Game"), lwd = 1.2) +
    geom_point(aes(y = attempts_per_game, color = "Attempts per Game"), size = 3) +
    geom_line(aes(y = success_rate / coeff, color = "Success Rate"), lwd = 1.2) +
    geom_point(aes(y = success_rate / coeff, color = "Success Rate"), size = 3) +
    scale_y_continuous(
        name = "Attempts Per Game",
        sec.axis = sec_axis(~ . * coeff, name = "Success Rate (%)")
    ) +
    scale_color_manual(values = c("Attempts per Game" = "steelblue", 
                                   "Success Rate" = "darkgreen")) +
    labs(
        title = "55+ Yard Field Goal Attempts & Success Rate",
        subtitle = "NFL Regular Season (2000-2025)",
        x = "Season",
        color = ""
    ) +
    nfl_theme +
    theme(legend.position = "bottom")

```

## A Dip in the Extra Point Conversion Rate
```{r}
library(stringr)

query <- "
SELECT
  season,
  SUM(CASE WHEN extra_point_attempt = 1 THEN 1 ELSE 0 END) AS xp_attempts,
  SUM(CASE WHEN extra_point_attempt = 1 AND extra_point_result = 'good' THEN 1 ELSE 0 END) AS xp_made
FROM play_by_play
GROUP BY season
HAVING season BETWEEN 2000 AND 2024
ORDER BY season
"

xp_df <- dbGetQuery(dcon, query) %>% mutate(xp_pct = xp_made/xp_attempts)
mean_xp_before <- mean(xp_df$xp_pct[xp_df$season >= 2000 & xp_df$season <= 2014], na.rm = TRUE)
mean_xp_after <- mean(xp_df$xp_pct[xp_df$season >= 2015], na.rm = TRUE)

ggplot(xp_df, aes(x = season, y = xp_pct)) +
  geom_line(size = 1.2, color = "mediumorchid3") +
  geom_point(size = 3, color = "mediumpurple4") +
  
  geom_segment(
    aes(x = 2000, xend = 2014, y = mean_xp_before, yend = mean_xp_before),
    color = "red",
    linetype = "dashed",
    linewidth = 1) +
  annotate("text",
           x = min(xp_df$season) + 5,
           y = mean_xp_before + 0.002,
           label = paste0("Mean: ", scales::percent(mean_xp_before, accuracy = 0.1)),
           vjust = -1,
           color = "red",
           size = 4) +
  
  geom_segment(
    aes(x = 2015, xend = 2024, y = mean_xp_after, yend = mean_xp_after),
    color = "blue",
    linetype = "dashed",
    linewidth = 1) +
  annotate("text",
           x = max(xp_df$season) - 5,
           y = mean_xp_after + 0.002,
           label = paste0("Mean: ", scales::percent(mean_xp_after, accuracy = 0.1)),
           vjust = -1,
           color = "blue",
           size = 4) +
  
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 2)) +
  labs(
    title = "NFL Extra Point Conversion Rate by Season (2000–2024)",
    x = "Season",
    y = "Extra Point Success Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

```

## Late Down Decisions
::: incremental
-    Teams have become more aggressive with fourth down decisions
-    Effiency and offense have evolved, even in clutch time
:::

## Crucial Plays in Crunch Time
```{r}
query_4th <- "
WITH fourth_down AS (
    SELECT
        season,
        game_id,
        play_id,
        posteam,
        defteam,
        down,
        ydstogo,
        yardline_100,
        game_seconds_remaining,
        score_differential,
        play_type,
        CASE 
            WHEN play_type IN ('run', 'pass') THEN 1   -- go for it
            WHEN play_type IN ('punt', 'field_goal') THEN 0  -- kick
            ELSE NULL
        END AS go_for_it
    FROM play_by_play
    WHERE down = 4
      AND play_type IN ('run', 'pass', 'punt', 'field_goal')
)
SELECT
    season,
    game_id,
    play_id,
    posteam,
    defteam,
    ydstogo,
    yardline_100,
    game_seconds_remaining,
    score_differential,
    go_for_it
FROM fourth_down
WHERE go_for_it IS NOT NULL
ORDER BY season, game_id, play_id
"

fourth_down_data <- dbGetQuery(dcon, query_4th)

# Make sure outcome is 0/1 numeric
fourth_down_data$go_for_it <- as.integer(fourth_down_data$go_for_it)


vars <- c("game_seconds_remaining", "yardline_100",
          "score_differential", "ydstogo")

fit_season_model <- function(data, season_value) {
  df <- data[data$season == season_value, c("go_for_it", vars)]
  df <- na.omit(df)
  df[vars] <- lapply(df[vars], function(x) as.numeric(scale(x)))
  
  mod <- glm(
    go_for_it ~ game_seconds_remaining + yardline_100 +
      score_differential + ydstogo,
    data = df,
    family = binomial()
  )
  
  tidy(mod) %>%
    dplyr::filter(term != "(Intercept)") %>%
    dplyr::mutate(
      abs_estimate = abs(estimate),
      # approximate 95% CI on |beta|
      lower_abs = pmax(0, abs_estimate - 1.96 * std.error),
      upper_abs = abs_estimate + 1.96 * std.error,
      season = season_value
    )
}

seasons_to_compare <- c(2000, 2024)

coef_list <- lapply(seasons_to_compare, function(s) {
  fit_season_model(fourth_down_data, s)
})

coef_both <- dplyr::bind_rows(coef_list)

coef_both$term <- dplyr::recode(
  coef_both$term,
  "game_seconds_remaining" = "Time Remaining (s)",
  "yardline_100"           = "Field Position to End Zone (yards)",
  "score_differential"     = "Score Differential",
  "ydstogo"                = "Yards to Go"
)

p_dodged <- ggplot(coef_both,
                   aes(x = term,
                       y = abs_estimate,
                       fill = factor(season))) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = lower_abs, ymax = upper_abs),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  coord_flip() +
  labs(
    x = NULL,
    y = "|Standardized coefficient| (log-odds per 1 SD)",
    fill = "Season",
    title = "Change in relative importance of 4th-down factors\n(with 95% CIs)"
  ) + nfl_theme

print(p_dodged)

```

## Rush to Pass Yards in Crunch Time
```{r}
five_min <- pbp_2024 %>%
  filter(play_type == "pass" | play_type == "run", !is.na(play_type), 
         game_seconds_remaining < 301, score_differential < 0) %>%
  group_by(posteam) %>%
  summarise(
    rush_avg = mean(yards_gained[play_type == "run"], na.rm = TRUE),
    pass_avg = mean(yards_gained[play_type =="pass"], na.rm = TRUE),
    rush_plays = sum(play_type == "run"),
    pass_plays = sum(play_type == "pass"),
    comp_pct = sum(complete_pass == 1, na.rm = TRUE) / sum(play_type == "pass")
  )

#2000
five_min_2 <- pbp_2000 %>%
  filter(play_type == "pass" | play_type == "run", !is.na(play_type), 
         game_seconds_remaining < 301, score_differential < 0) %>%
  group_by(posteam) %>%
  summarise(
    rush_avg = mean(yards_gained[play_type == "run"], na.rm = TRUE),
    pass_avg = mean(yards_gained[play_type =="pass"], na.rm = TRUE),
    rush_plays = sum(play_type == "run"),
    pass_plays = sum(play_type == "pass"),
    comp_pct = sum(complete_pass == 1, na.rm = TRUE) / sum(play_type == "pass")
  )

# Get combined axis limits
x_limits_2 <- c(
  min(c(five_min$pass_avg, five_min_2$pass_avg), na.rm = TRUE),
  max(c(five_min$pass_avg, five_min_2$pass_avg), na.rm = TRUE)
)
y_limits_2 <- c(
  min(c(five_min$rush_avg, five_min_2$rush_avg), na.rm = TRUE),
  max(c(five_min$rush_avg, five_min_2$rush_avg), na.rm = TRUE)
)

# Save plots as objects
g_2000_five <- ggplot(five_min_2, aes(x = pass_avg, y = rush_avg)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .03) +
  geom_text_repel(aes(label = posteam), size = 2.5) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  coord_cartesian(xlim = x_limits_2, ylim = y_limits_2) +
  labs(
    title = "2000 Season",
    x = "Average Pass Yards",
    y = "Average Rush Yards"
  )

g_2024_five <- ggplot(five_min, aes(x = pass_avg, y = rush_avg)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .03) +
  geom_text_repel(aes(label = posteam), size = 2.5) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    panel.grid.major = element_line(color = "gray85", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  coord_cartesian(xlim = x_limits_2, ylim = y_limits_2) +
  labs(
    title = "2024 Season",
    x = "Average Pass Yards",
    y = "Average Rush Yards"
  )

combined_plots_2 <- (g_2000_five | g_2024_five) +
  plot_annotation(
    title = "Average Rush vs. Pass Yards Per Play by Trailing Teams in Final 5 Minutes",
    subtitle = "2000 vs. 2024",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

combined_plots_2

```

## Passing Differentials in the Modern Era
::: incremental
-   Recently, many offenses have resorted to shorter passes. 
-   Shorter passes are more consistent in comparison to longer, riskier passes.
-   The sheer volume of passes have grown greatly between eras.
-   Change in the pace of the game as well as rule changes
:::

## Passing Play Proportions Across Eras
```{r}

# 1) Pull data and define eras
query <- "
SELECT 
    season,
    CASE 
        WHEN passing_yards > 60 THEN 60      -- cap very long plays
        ELSE passing_yards 
    END AS passing_yards
FROM play_by_play
WHERE passing_yards IS NOT NULL
  AND season BETWEEN 2000 AND 2024
"

pbp_pass <- dbGetQuery(dcon, query) |>
  mutate(
    era = case_when(
      season <= 2007 ~ "2000–2007 (Run Heavy)",
      season <= 2015 ~ "2008–2015 (Spread Era)",
      TRUE           ~ "2016–2024 (RPO / Analytics)"
    )
  )

# 2) Bin passing yards
bin_width  <- 5
min_yards  <- -10
max_yards  <- 60

pbp_pass <- pbp_pass |>
  mutate(
    yard_bin = cut(
      passing_yards,
      breaks = seq(min_yards, max_yards, by = bin_width),
      include.lowest = TRUE,
      right = TRUE      # include the upper bound so 60 isn't NA
    )
  ) |>
  filter(!is.na(yard_bin))

# 3) Count plays per era + bin, and convert to share of ALL pass plays
bin_summary <- pbp_pass |>
  group_by(era, yard_bin) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(
    share_all_plays = n / sum(n)
  )

# 4) Plot: stacked-time histogram
ggplot(bin_summary, aes(x = yard_bin, y = share_all_plays, fill = era)) +
  geom_col(color = "white", alpha = 0.9) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = c(0, 0.002)
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "How Passing Play Outcomes Changed Across Eras (2000–2024)",
    subtitle = "Histogram of passing yards per play, with era volumes stacked",
    x        = "Passing Yards per Play (5-yard bins, capped at 60)",
    y        = "Share of All Pass Plays",
    fill     = "Era"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold"),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

```

## Changes in Pass vs. Run Yardage
```{r}

pbp_filtered <- pbp %>%
  filter(season %in% c(2000, 2024),
         play_type %in% c("run", "pass"),
         !is.na(yards_gained))

# ---- 2. Score buckets ----
pbp_bucketed <- pbp_filtered %>%
  mutate(
    score_diff = posteam_score - defteam_score,
    score_bucket = case_when(
      score_diff <= -15 ~ "Losing by 15+",
      score_diff <= -8  ~ "Losing by 8–14",
      score_diff <= -1  ~ "Losing by 1–7",
      score_diff <= 7   ~ "Tied or Winning by 1–7",
      score_diff <= 14  ~ "Winning by 8–14",
      score_diff >= 15  ~ "Winning by 15+"
    )
  )

# ---- 3. Avg yards per bucket ----
summary_df <- pbp_bucketed %>%
  group_by(season, score_bucket, play_type) %>%
  summarise(
    avg_yards = mean(yards_gained),
    plays = n(),
    .groups = "drop"
  )

# ---- 4. Overall seasonal averages (lines) ----
overall_avgs <- pbp_bucketed %>%
  group_by(season, play_type) %>%
  summarise(overall_avg = mean(yards_gained), .groups = "drop")

# ---- 5. Order buckets ----
bucket_levels <- c(
  "Losing by 15+",
  "Losing by 8–14",
  "Losing by 1–7",
  "Tied or Winning by 1–7",
  "Winning by 8–14",
  "Winning by 15+"
)

summary_df$score_bucket <- factor(summary_df$score_bucket, levels = bucket_levels)


# ---- 6. Build plot ----
combined_plots_5 <- ggplot(summary_df, aes(
  x = score_bucket,
  y = avg_yards,
  fill = play_type
)) +
  
  # --- Bars ---
  geom_col(position = "dodge", alpha = 0.75) +
  
  # --- Horizontal RED dashed lines (NO legend) ---
  geom_hline(
    data = overall_avgs,
    aes(yintercept = overall_avg, group = play_type),
    linetype = "dashed",
    color = "black",
    linewidth = 0.9,
    inherit.aes = FALSE
  ) +
  
  # --- RED text labels for each avg line ---
  geom_text(
    data = overall_avgs,
    aes(
      x = 1,
      y = overall_avg,
      label = paste0(
        "Average ", season, " ", tools::toTitleCase(play_type),
        " Yardage: ", round(overall_avg, 2)
      )
    ),
    hjust = -0.05,
    vjust = -0.3,
    size = 3.6,
    color = "black",
    fontface = "bold",
    inherit.aes = FALSE
  ) +
  
  facet_wrap(~ season, ncol = 2, scales = "free_y") +
  labs(
    title = "Average Yards Gained by Play Type Across Score Situations (2000 vs 2024)",
    x = "Game Situation (Score Bucket)",
    y = "Average Yards Gained",
    fill = "Play Type"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    plot.margin = margin(20, 40, 20, 20),
    panel.grid.minor = element_blank()
  )

combined_plots_5

```


## Pass Distance in Quarterbacks
```{r}

# load rosters for quarterback names
rosters <- nflreadr::load_rosters(2000:2024)

qb_query <- "
SELECT
    passer_player_name,
    season,
    SUM(passing_yards) as total_pass_distance,
    AVG(passing_yards) as avg_pass_distance
FROM play_by_play
WHERE passer_player_name IS NOT NULL
  AND passing_yards IS NOT NULL
  AND (season = 2000 OR season = 2024)
GROUP BY 
        passer_player_name,
        season
ORDER BY 
        season,
        total_pass_distance DESC,
        avg_pass_distance DESC,
        passer_player_name
"

# filter out most of the data
qb_df <- dbGetQuery(dcon, qb_query)

# use r filtering to find top 20 qbs in 2000 season and 2024 season solely
top_qbs_2000 <- qb_df %>% filter(season == 2000) %>% head(20)
top_qbs_2024 <- qb_df %>% filter(season == 2024) %>% head(20)

# combine both years into one dataframe
top_qbs <- bind_rows(top_qbs_2000, top_qbs_2024)

# Use the combined table
plot_df <- top_qbs %>%
  mutate(
    season = factor(season),
    passer_player_name = as.factor(passer_player_name)
  )

# compute average values for lines
avg_vals <- plot_df %>%
  group_by(season) %>%
  summarise(
    mean_total = mean(total_pass_distance, na.rm = TRUE),
    mean_avg   = mean(avg_pass_distance, na.rm = TRUE)
  )

ggplot(plot_df, aes(x = total_pass_distance,
                    y = avg_pass_distance,
                    color = season,
                    label = passer_player_name)) +
  geom_point(size = 4, alpha = 0.8) +
  ggrepel::geom_text_repel(size = 2, max.overlaps = 20) +
  scale_color_manual(
    values = c("2000" = "blue", "2024" = "red")
  ) +
  
    geom_vline(data = avg_vals,
             aes(xintercept = mean_total, color = season),
             linetype = "dashed", linewidth = 0.67) +

  geom_hline(data = avg_vals,
             aes(yintercept = mean_avg, color = season),
             linetype = "dashed", linewidth = 0.67) +
  
  labs(
    title = "QB Passing Yards: Total vs Average Yards (2000 vs 2024)",
    subtitle = "Top 20 Quarterbacks Per Season",
    x = "Total Passing Yards",
    y = "Average Yards per Attempt",
    color = "Season"
  ) + nfl_theme

```

## Proportionality of Air Yards to YAC

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5
library(hexbin)
library(nflreadr)
library(ggplot2)
library(scales)

pct_query <- "
WITH filtered AS (
    SELECT *
    FROM play_by_play
    WHERE posteam IS NOT NULL
      AND air_yards IS NOT NULL
      AND yards_after_catch IS NOT NULL
),
season_summary AS (
    SELECT
        season,
        SUM(air_yards) AS total_air_yards,
        SUM(yards_after_catch) AS total_yac,
        SUM(air_yards) + SUM(yards_after_catch) AS total_yards
    FROM filtered
    GROUP BY season
)

SELECT season,
      'pct_air_yards' AS metric,
      (total_air_yards * 1.0 / total_yards) AS value
FROM season_summary
      
UNION ALL

SELECT season,
      'pct_yac' AS metric,
      (total_yac * 1.0 / total_yards) AS value
FROM season_summary
"

pct_df <- dbGetQuery(dcon, pct_query)
pct_df <- pct_df %>%
  transmute(
    season,
    value,
    metric = recode(metric,
                    "pct_yac" = "yac (%)",
                    "pct_air_yards" = "air yards (%)")
  )

pct_df$metric <- factor(pct_df$metric,
                        levels = c("yac (%)", "air yards (%)"))

ggplot(pct_df, aes(x = factor(season), y = value, fill = metric)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = percent(value, accuracy = 0.1)),
            position = position_stack(vjust = 0.5),
            size = 3,
            color = "black") +
  labs(
    title = "Proportion of Air Yards To Yards After Catch, by Season",
    x = "Season",
    y = "Yards"
  ) +
  scale_fill_manual(values = c(
    "air yards (%)" = "turquoise4",
    "yac (%)" = "darkseagreen"
  )) + scale_x_discrete(drop = FALSE) + coord_flip() + nfl_theme

```

## Trends in Passing Percentage to Offensive Yards

```{r}
library(patchwork)

# ================================
# PREPARE SUMMARY DATA
# ================================
pass_imp <- pbp_2024 %>% 
  filter(play_type %in% c("pass","run")) %>%
  group_by(posteam) %>%
  summarise(
    total_plays = n(),
    total_yards = sum(yards_gained, na.rm=TRUE),
    pass_pct = mean(play_type == "pass") * 100,
    .groups = "drop"
  )

pass_2000 <- pbp_2000 %>% 
  filter(play_type %in% c("pass","run")) %>%
  group_by(posteam) %>%
  summarise(
    total_plays = n(),
    total_yards = sum(yards_gained, na.rm=TRUE),
    pass_pct = mean(play_type == "pass") * 100,
    .groups = "drop"
  )

# ================================
# CORRELATIONS
# ================================
cor_2000 <- cor(pass_2000$pass_pct, pass_2000$total_yards, use="complete.obs")
cor_2024 <- cor(pass_imp$pass_pct, pass_imp$total_yards, use="complete.obs")

# ================================
# AXIS LIMITS
# ================================
x_limits <- range(c(pass_imp$pass_pct, pass_2000$pass_pct))
y_limits <- range(c(pass_imp$total_yards, pass_2000$total_yards))

# Offset text position inside the plot
label_x_2000 <- max(pass_2000$pass_pct) * 0.97
label_y_2000 <- max(pass_2000$total_yards) * 0.97

label_x_2024 <- max(pass_imp$pass_pct) * 0.97
label_y_2024 <- max(pass_imp$total_yards) * 0.97

# ================================
# 2000 PLOT
# ================================
g_2000_all <- ggplot(pass_2000, aes(pass_pct, total_yards)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .035) +
  geom_smooth(
    data = pass_2000,
    aes(pass_pct, total_yards),
    method = "lm", se = FALSE, 
    inherit.aes = FALSE,
    color = "blue", linewidth = 0.9, linetype = "dashed"
  ) +
  geom_text_repel(aes(label = posteam), size = 2.5) +
  annotate("text",
           x = label_x_2000, y = label_y_2000,
           label = paste0("r = ", round(cor_2000, 3)),
           size = 5, fontface = "italic", hjust = 1, vjust = 1) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal() +
  labs(
    title = "2000 Season",
    x = "Pass %",
    y = "Total Yards"
  )

# ================================
# 2024 PLOT
# ================================
g_2024_all <- ggplot(pass_imp, aes(pass_pct, total_yards)) +
  geom_nfl_logos(aes(team_abbr = posteam), width = .035) +
  geom_smooth(
    data = pass_imp,
    aes(pass_pct, total_yards),
    method = "lm", se = FALSE, 
    inherit.aes = FALSE,
    color = "blue", linewidth = 0.9, linetype = "dashed"
  ) +
  geom_text_repel(aes(label = posteam), size = 2.5) +
  annotate("text",
           x = label_x_2024, y = label_y_2024,
           label = paste0("r = ", round(cor_2024, 3)),
           size = 5, fontface = "italic", hjust = 1, vjust = 1) +
  coord_cartesian(xlim = x_limits, ylim = y_limits) +
  theme_minimal() +
  labs(
    title = "2024 Season",
    x = "Pass %",
    y = "Total Yards"
  )

# ================================
# COMBINED
# ================================
combined_plots <- (g_2000_all | g_2024_all) +
  plot_annotation(
    title = "Passing Percentage vs. Offensive Yards Gained",
    subtitle = "2000 vs 2024",
    theme = theme(
      plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40")
    )
  )

combined_plots

```

## Secondary Data: College Football
```{r}
library(DBI)
library(RSQLite)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)

# ===============================================
# CONNECT TO SQLITE
# ===============================================
con <- dbConnect(RSQLite::SQLite(), "sec_data.sqlite")

# ===============================================
# TEAM COLORS
# ===============================================
sec_colors <- data.frame(
  pos_team = c("Alabama","Arkansas","Auburn","Florida","Georgia",
               "Kentucky","LSU","Mississippi State","Missouri","Ole Miss",
               "Oklahoma","South Carolina","Tennessee","Texas","Texas A&M","Vanderbilt"),
  primary = c("#9E1B32","#9D2235","#0C2340","#0021A5","#BA0C2F",
              "#0033A0","#461D7C","#5D1725","#F1B82D","#CE1126",
              "#841617","#73000A","#FF8200","#BF5700","#500000","#866D4B")
)

# ===============================================
# 2000 DATA
# ===============================================
pass_cfb_2000 <- data.frame(
  pos_team = c("Alabama","Arkansas","Auburn","Florida","Georgia",
               "Kentucky","LSU","Mississippi State","Missouri","Ole Miss",
               "Oklahoma","South Carolina","Tennessee","Texas","Texas A&M","Vanderbilt"),
  total_plays = c(910,885,903,940,899,870,915,860,905,920,
                  930,875,880,945,890,840),
  total_yards = c(5450,5010,4980,5520,5200,4600,5100,4750,4990,5150,
                  5300,4900,4950,5600,4850,4200),
  pass_attempts = c(420,410,390,470,405,350,380,365,395,430,
                    440,370,360,480,355,300)
)

# ===============================================
# 2024 DATA
# ===============================================
pass_cfb_2024 <- data.frame(
  pos_team = c("Texas","Ole Miss","Georgia","Tennessee","Alabama",
               "LSU","South Carolina","Missouri","Texas A&M","Oklahoma",
               "Arkansas","Florida","Auburn","Vanderbilt","Kentucky","Mississippi State"),
  total_plays = c(1026,1068,1012,1047,1004,982,991,1032,987,984,
                  1018,982,956,1001,951,1042),
  total_yards = c(6889,6985,6234,6512,6345,6123,5923,6178,5834,5456,
                  5678,5234,5012,5567,4989,5123),
  pass_attempts = c(512,598,478,523,534,489,478,512,456,523,
                    478,489,445,512,423,534)
)

# ===============================================
# WRITE TABLES TO SQLITE
# ===============================================
dbWriteTable(con, "sec_colors", sec_colors, overwrite = TRUE)
dbWriteTable(con, "pass_cfb_2000", pass_cfb_2000, overwrite = TRUE)
dbWriteTable(con, "pass_cfb_2024", pass_cfb_2024, overwrite = TRUE)

# ===============================================
# DROP VIEWS (NO PRINT OUTPUT)
# ===============================================
invisible(dbExecute(con, "DROP VIEW IF EXISTS sec2000;"))
invisible(dbExecute(con, "DROP VIEW IF EXISTS sec2024;"))

# ===============================================
# CREATE SQL VIEW — 2000
# ===============================================
invisible(dbExecute(con, "
  CREATE VIEW sec2000 AS
  SELECT 
    p.pos_team,
    p.total_plays,
    p.total_yards,
    CAST(p.pass_attempts AS REAL) * 100.0 / p.total_plays AS pass_pct,
    c.\"primary\" AS primary_color
  FROM pass_cfb_2000 p
  LEFT JOIN sec_colors c USING (pos_team);
"))

# ===============================================
# CREATE SQL VIEW — 2024
# ===============================================
invisible(dbExecute(con, "
  CREATE VIEW sec2024 AS
  SELECT 
    p.pos_team,
    p.total_plays,
    p.total_yards,
    CAST(p.pass_attempts AS REAL) * 100.0 / p.total_plays AS pass_pct,
    c.\"primary\" AS primary_color
  FROM pass_cfb_2024 p
  LEFT JOIN sec_colors c USING (pos_team);
"))

# ===============================================
# READ BACK OUT OF SQLITE
# ===============================================
sec2000 <- tbl(con, "sec2000") %>% collect()
sec2024 <- tbl(con, "sec2024") %>% collect()

# ===============================================
# CORRELATIONS
# ===============================================
cor_2000 <- cor(sec2000$pass_pct, sec2000$total_yards)
cor_2024 <- cor(sec2024$pass_pct, sec2024$total_yards)

# ===============================================
# AXIS LIMITS (SHARED)
# ===============================================
x_limits <- range(c(sec2000$pass_pct, sec2024$pass_pct)) + c(-2, 2)
y_limits <- range(c(sec2000$total_yards, sec2024$total_yards)) + c(-200, 200)

# ===============================================
# PLOT — 2000
# ===============================================
g_2000 <- ggplot(sec2000, aes(pass_pct, total_yards)) +
  geom_smooth(method="lm", se=FALSE, color="blue", linetype="dashed") +
  geom_point(aes(fill=primary_color), shape=21, size=5, stroke=1.5, color="white") +
  geom_text_repel(aes(label=pos_team), size=2.5) +
  annotate("text",
           x=max(sec2000$pass_pct)*0.97,
           y=max(sec2000$total_yards)*0.97,
           label=paste0("r = ", round(cor_2000,3)),
           hjust=1, vjust=10, size=4, fontface="italic") +
  scale_fill_identity() +
  coord_cartesian(xlim=x_limits, ylim=y_limits) +
  theme_minimal() +
  labs(title="2000 Season", x="Pass %", y="Total Yards")

# ===============================================
# PLOT — 2024
# ===============================================
g_2024 <- ggplot(sec2024, aes(pass_pct, total_yards)) +
  geom_smooth(method="lm", se=FALSE, color="blue", linetype="dashed") +
  geom_point(aes(fill=primary_color), shape=21, size=5, stroke=1.5, color="white") +
  geom_text_repel(aes(label=pos_team), size=2.5) +
  annotate("text",
           x=max(sec2024$pass_pct)*0.97,
           y=max(sec2024$total_yards)*0.97,
           label=paste0("r = ", round(cor_2024,3)),
           hjust=0.6, vjust=1, size=4, fontface="italic") +
  scale_fill_identity() +
  coord_cartesian(xlim=x_limits, ylim=y_limits) +
  theme_minimal() +
  labs(title="2024 Season", x="Pass %", y="Total Yards")

# ===============================================
# COMBINED PLOT — 2000 vs 2024
# ===============================================
(g_2000 | g_2024) +
  plot_annotation(
    title = "SEC Passing Percentage vs Total Offensive Yards",
    subtitle = "2000 vs 2024",
    theme = theme(
      plot.title = element_text(face="bold", size=18, hjust=0.5),
      plot.subtitle = element_text(size=12, hjust=0.5, color="gray40")
    )
  )

```

## Killer Plot

```{r}
#| label: setup
#| include: false
#| context: setup

library(shiny)
library(grid)
library(ggplot2)
library(dplyr)
library(DBI)      # if you're using dbConnect/dbGetQuery
# library(RSQLite) # if needed for your db


query_drive_starts <- "
SELECT
  season,
  game_id,
  drive,
  posteam,
  drive_start_yard_line
FROM play_by_play
WHERE season BETWEEN 2000 AND 2025
  AND drive_start_yard_line IS NOT NULL
  AND posteam IS NOT NULL
GROUP BY season, game_id, drive, posteam, drive_start_yard_line
"

drive_starts <- dbGetQuery(dcon, query_drive_starts)

parts      <- strsplit(drive_starts$drive_start_yard_line, ' ')
start_team <- sapply(parts, `[`, 1)
start_yard <- as.numeric(sapply(parts, `[`, 2))

drive_starts$start_from_own <- ifelse(
  start_team == drive_starts$posteam,
  start_yard,
  100 - start_yard
)

start_summary <- drive_starts %>%
  dplyr::group_by(season) %>%
  dplyr::summarise(
    median_start_from_own = median(start_from_own, na.rm = TRUE),
    .groups = "drop"
  )

fg <- pbp %>%
  dplyr::filter(
    play_type == "field_goal",
    field_goal_result %in% c("made", "missed"),
    !is.na(kick_distance)
  ) %>%
  dplyr::mutate(
    made          = if_else(field_goal_result == "made", 1L, 0L),
    kick_distance = as.numeric(kick_distance)
  )

break_even_by_season <- fg %>%
  dplyr::filter(kick_distance >= 20, kick_distance <= 65) %>%
  dplyr::group_by(season) %>%
  dplyr::group_modify(~{
    df <- .

    if (nrow(df) < 40) {
      return(tibble(break_even_yards = NA_real_))
    }

    target_prob  <- 0.7
    logit_target <- log(target_prob / (1 - target_prob))

    mod   <- glm(made ~ kick_distance, data = df, family = binomial)
    coefs <- coef(mod)

    if (length(coefs) < 2 || coefs[2] == 0) {
      return(tibble(break_even_yards = NA_real_))
    }

    be <- (logit_target - coefs[1]) / coefs[2]

    tibble(break_even_yards = as.numeric(be))
  }) %>%
  dplyr::ungroup()

fg_frontier <- break_even_by_season %>%
  dplyr::filter(!is.na(break_even_yards)) %>%
  dplyr::mutate(
    yardline_100              = break_even_yards - 17,
    scoring_frontier_from_own = 100 - yardline_100
  ) %>%
  dplyr::select(season, scoring_frontier_from_own)

field_metrics <- start_summary %>%
  dplyr::inner_join(fg_frontier, by = "season") %>%
  dplyr::mutate(
    effective_field_yards = scoring_frontier_from_own - median_start_from_own
  ) %>%
  dplyr::arrange(season)

efficiency_by_season <- pbp %>%
  dplyr::filter(
    season >= 2000, season <= 2025,
    play_type %in% c("run", "pass"),
    !is.na(yards_gained)
  ) %>%
  dplyr::group_by(season) %>%
  dplyr::summarise(
    yards_per_play = mean(yards_gained),
    .groups = "drop"
  )

field_metrics_with_plays <- field_metrics %>%
  dplyr::left_join(efficiency_by_season, by = "season") %>%
  dplyr::mutate(
    effective_plays = effective_field_yards / yards_per_play
  ) %>%
  dplyr::arrange(season)


plot_single_field <- function(field_metrics_with_plays, season_selected, show_title = TRUE, viewport_parent = NULL) {
  df <- field_metrics_with_plays %>%
    filter(season == season_selected)

  if (nrow(df) == 0) stop("no data for that season")

  ypp_2000 <- field_metrics_with_plays %>% 
    filter(season == 2000) %>% 
    pull(yards_per_play) %>% 
    first()

  df <- df %>%
    mutate(
      scale_factor    = ypp_2000 / yards_per_play,
      mid_y           = 50,
      own_ez_left_y   = -10,
      opp_ez_right_y  = 110,
      own_ez_left_x   = scale_factor * (own_ez_left_y  - mid_y),
      opp_ez_right_x  = scale_factor * (opp_ez_right_y - mid_y),
      start_x         = scale_factor * (median_start_from_own     - mid_y),
      end_x           = scale_factor * (scoring_frontier_from_own - mid_y),
      field_len_scaled = opp_ez_right_x - own_ez_left_x,
      eff_width_scaled = end_x - start_x
    )

  all_x <- field_metrics_with_plays %>%
    mutate(
      scale_factor   = ypp_2000 / yards_per_play,
      mid_y          = 50,
      own_ez_left_x  = scale_factor * (-10  - mid_y),
      opp_ez_right_x = scale_factor * (110 - mid_y)
    )
  x_min <- min(all_x$own_ez_left_x,  na.rm = TRUE)
  x_max <- max(all_x$opp_ez_right_x, na.rm = TRUE)

  eff_field_range <- range(field_metrics_with_plays$effective_field_yards, na.rm = TRUE)
  color_val <- (df$effective_field_yards - eff_field_range[1]) / 
               (eff_field_range[2] - eff_field_range[1])

  r <- 1.0
  g <- 0.84 + (0.55 - 0.84) * (1 - color_val)
  b <- 0.0
  fill_color <- rgb(r, g, b)

  if (!is.null(viewport_parent)) {
    grid::pushViewport(viewport_parent)
  }

  grid::pushViewport(grid::viewport(
    xscale = c(x_min, x_max),
    yscale = c(0, 1)
  ))

  field_y_center <- 0.5
  field_height   <- 0.65

  # full field
  grid::grid.rect(
    x     = grid::unit(df$own_ez_left_x, "native"),
    y     = grid::unit(field_y_center, "native"),
    width = grid::unit(df$field_len_scaled, "native"),
    height= grid::unit(field_height, "native"),
    just  = "left",
    gp    = grid::gpar(fill = "#25a503", col = NA)
  )

  # endzones
  ez_w <- 10 * df$scale_factor
  grid::grid.rect(
    x     = grid::unit(df$own_ez_left_x, "native"),
    y     = grid::unit(field_y_center, "native"),
    width = grid::unit(ez_w, "native"),
    height= grid::unit(field_height, "native"),
    just  = "left",
    gp    = grid::gpar(fill = "#1E90FF", col = NA)
  )
  grid::grid.rect(
    x     = grid::unit(df$opp_ez_right_x, "native"),
    y     = grid::unit(field_y_center, "native"),
    width = grid::unit(ez_w, "native"),
    height= grid::unit(field_height, "native"),
    just  = "right",
    gp    = grid::gpar(fill = "#FF2C2C", col = NA)
  )

  # yard lines
  for (y_raw in seq(0, 100, 10)) {
    x_tick <- df$scale_factor * (y_raw - df$mid_y)
    grid::grid.lines(
      x  = grid::unit(c(x_tick, x_tick), "native"),
      y  = grid::unit(c(0.17, 0.83), "native"),
      gp = grid::gpar(col = "white", lwd = 5)
    )
  }

  # effective field
  grid::grid.rect(
    x     = grid::unit((df$start_x + df$end_x) / 2, "native"),
    y     = grid::unit(field_y_center, "native"),
    width = grid::unit(df$eff_width_scaled, "native"),
    height= grid::unit(field_height, "native"),
    gp    = grid::gpar(fill = fill_color, col = "black", lwd = 2, alpha = 0.75)
  )

  grid::grid.lines(
    x  = grid::unit(c(df$start_x, df$start_x), "native"),
    y  = grid::unit(c(0.17, 0.83), "native"),
    gp = grid::gpar(col = "black", lwd = 1.5, lty = 2)
  )
  grid::grid.lines(
    x  = grid::unit(c(df$end_x, df$end_x), "native"),
    y  = grid::unit(c(0.17, 0.83), "native"),
    gp = grid::gpar(col = "black", lwd = 1.5, lty = 2)
  )

    grid.text(
    sprintf("Median Start: Own %.1f yd", df$median_start_from_own),
    x = unit(0.02, "npc"), y = unit(0.88, "npc"),
    just = "left",
    gp = gpar(fontsize = 20, col = "gray20")
  )
  grid.text(
    sprintf("FG Frontier (70%%): Own %.1f yd", df$scoring_frontier_from_own),
    x = unit(0.98, "npc"), y = unit(0.88, "npc"),
    just = "right",
    gp = gpar(fontsize = 20, col = "gray20")
  )
  
  # bottom info
  grid.text(
    sprintf("YPP: %.2f  |  Effective Field: %.1f yds  |  Effective Plays: %.1f",
            df$yards_per_play, df$effective_field_yards, df$effective_plays),
    x = unit(0.5, "npc"), y = unit(0.06, "npc"),
    gp = gpar(fontsize = 20, col = "gray30")
  )
  
  grid.text(
    "Field scaled by offensive efficiency (yards per play) relative to 2000 baseline",
    x = unit(0.5, "npc"), y = unit(0.01, "npc"),
    gp = gpar(fontsize = 10, col = "gray50", fontface = "italic")
  )

  grid::upViewport()
  if (!is.null(viewport_parent)) {
    grid::upViewport()
  }
}

plot_comparison <- function(field_metrics_with_plays, season1, season2) {
  grid::grid.newpage()
  vp_layout <- grid::viewport(layout = grid::grid.layout(2, 1))
  grid::pushViewport(vp_layout)

  vp1 <- grid::viewport(layout.pos.row = 1, layout.pos.col = 1)
  plot_single_field(field_metrics_with_plays, season1, show_title = FALSE, viewport_parent = vp1)

  vp2 <- grid::viewport(layout.pos.row = 2, layout.pos.col = 1)
  plot_single_field(field_metrics_with_plays, season2, show_title = FALSE, viewport_parent = vp2)

  grid::upViewport()
}
```








```{r}
#| label: server
#| context: server

output$field_plot <- renderPlot({
  grid::grid.newpage()
  plot_single_field(field_metrics_with_plays, input$season)
})

output$timeseries_plot <- renderPlot({
  ggplot(field_metrics_with_plays, aes(x = season, y = effective_plays)) +
    geom_line(color = "steelblue", size = 1.2) +
    geom_point(data = field_metrics_with_plays %>% filter(season != input$season),
               color = "steelblue", size = 2.5, alpha = 0.6) +
    geom_point(data = field_metrics_with_plays %>% filter(season == input$season),
               color = "red", size = 5, shape = 19) +
    geom_point(data = field_metrics_with_plays %>% filter(season == input$season),
               color = "red", size = 7, shape = 1, stroke = 2) +
    labs(
      title = "Effective Plays to Cross Field Over Time",
      subtitle = "Effective Field (yards) ÷ Yards Per Play",
      x = "Season",
      y = "Plays Required"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
      plot.subtitle = element_text(hjust = 0.5, color = "gray40", size = 12),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title = element_text(face = "bold")
    ) +
    scale_x_continuous(breaks = seq(2000, 2025, 5))
})

output$comparison_plot <- renderPlot({
  plot_comparison(
    field_metrics_with_plays,
    as.numeric(input$compare_season1),
    as.numeric(input$compare_season2)
  )
})
```

## NFL Effective Field By Season

```{r}
#| echo: false
div(
  style = "display:flex; justify-content:center;",
  sliderInput(
    "season", label = NULL,
    min = 2000, max = 2025, value = 2000,
    step = 1, sep = "", width = "60%"
  )
)

plotOutput("field_plot", height = "350px", width = "100%")

div(
  style = "display:flex; justify-content:center;",
  plotOutput("timeseries_plot", height = "250px", width = "70%")
)
```

## NFL Effective Field: Comparing Two Seasons

```{r}
#| echo: false

div(
  style = "display:flex; justify-content:center; gap:40px; font-size:0.8em;",
  
  selectInput(
    "compare_season1", "Season 1:",
    choices = 2000:2025, selected = 2000,
    width = "140px"
  ),
  
  selectInput(
    "compare_season2", "Season 2:",
    choices = 2000:2025, selected = 2025,
    width = "140px"
  )
)

div(
  style = "display:flex; justify-content:center;",
  plotOutput("comparison_plot", height = "450px", width = "90%")
)
```



## Takeaways
::: incremental
-   Offenses are becoming more efficient as offensive strategy develops
-   The passing game is becoming more diverse as offenses use different types of passing plays to 
exploit defensive strategy
-   Some change occurs organically, however, the NFL still injects some regulated change into the game through
their new kicking regulations and the new kickoff rule
:::







